---
title: "Bioeconomy Trends Assessment"
author: "Sarah Mubareka"
reviewer: "--"
date: "25/11/2021"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using visdat}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo = FALSE, include = FALSE}

knitr::opts_chunk$set(fig.width = 7,
                      fig.height = 4)



library(purrr)
library(tidyverse)
library(dplyr)
library(COINr)
library(naniar)
```
## Pre-processing indicators 

The code below shows the procedure for creating the composite view of trends of the EU Bioeconomy's main objectives. The approach is a hierarchical one, with treatment of indicators and aggregation to key component and normative criteria levels and ultimately to the Strategy Objectives level.

We start by exploring and cleaning the raw indicators and filling in data for the alternate years (some indicators report every other year only).

### Read all indicators csv file 
The indicators were bulk downloaded from the system internally. This is not yet possible from the Internet but is planned. One of the attributes downloaded is the Green Deal Key. Indicators may have many Green Deal priorities associated. This results in multiple entries per indicator. 
```{r}

merged_GD<- read.csv("all_indicators_Nov2021.csv")

#REMOVE GD COLUMN FOR NOW 
merged<-merged_GD %>%
  select(-gd_id)

```

### Check available years in data 
We explore the years available for the data
```{r}
yearminmax<-merged %>%
    group_by(indicator_id) %>%
  mutate(minyr = min(time),maxyr=max(time)) %>%
  distinct(indicator_id, minyr, maxyr) %>%
  arrange(indicator_id)
print(yearminmax)
```


### Delete unwanted rows
We know there are some redundant indicators (e.g. sum of parts) and we want to remove these. We also want to remove indicators that are from before 2008.Since we want to keep track of the rows deleted, these are listed one by one so they may be changed if data improves over time.
```{r}
merged_sub<-merged %>%
  subset(type!="VEG_share") %>%
  subset(type!="ANI_share") %>%
  subset(type!="TOTAL") %>%
  subset(type!="all_stages") %>%
  subset(type!="all_categories") %>%
  subset(type!="felling_rates") %>%
  subset(type!="TPA_PC")%>%
  subset(indicator_id!="2.3.a.2")%>%   #this is the only indicator under provisional services (RW removals)
  subset(time>2007)%>%
  subset(time<2019)%>%
  subset(geo_code!="27")%>%
  subset(geo_code!="37")%>%
  subset(geo_code!="UK")%>%
  subset(geo_code!="EU28")
  #subset(geo_code!="EU_V")%>%
 # subset(geo_code!="EU27_2020")

```

### Computing indicators at EU-27_2020 level  
To complete data coverage for all indicators we use values from the closest available year. Furthermore, we wish to compute the mean for the EU.
```{r}
#impute and retain years 2012 and 2017 only
merged_subf_MS <-merged_sub %>%
    distinct(indicator_id, type, geo_code,time, .keep_all=TRUE) %>%
    group_by(geo_code, type, indicator_id) %>% 
    complete(time = full_seq(2010:2018, 1)) %>% 
    fill(value, .direction="downup") %>%
  subset(time==2012 | time==2017)
```

Not all indicators have a ready-computed Eu-27 or EU-V level value. We have to find them and fill those. Furthermore, the emissions indicators are a strange animal in the dataset, first because the more the negative the better (i.e. they are sinks), but other indicators also have this characteristic (negative directionality) but mainly because the sinks start out as negative. Therefore they have to brought o a positive scale prior to further processing.

```{r}
#see all values in the geography column
allgeo<-as.data.frame(unique(merged_subf_MS$geo_code))
#Merge dataframes of EU calculated and the original, giving priority to the pre-calculated original
temp<- merged_subf_MS %>%
  group_by(indicator_id, type, time)%>%
  summarise_at(vars(value), funs(mean(., na.rm=TRUE))) %>%
  mutate (geo_code = "EU27_2020")

merged_EU_MS<-merged_subf_MS%>%
  merge(temp,all=TRUE) %>%
  distinct(indicator_id,geo_code,type,time, .keep_all=TRUE )%>%
  mutate(abs(value)) %>%
  select(-value) %>%
  rename(value="abs(value)")


```


We selected the period between the years 2012 and 2017 for the analysis because this period contains data for all indicators. We apply the approach used by ESTAT for the computation of trends towards Sustainable Development Goals (https://ec.europa.eu/eurostat/web/sdi) . They apply a CAGR (Compound Annual Growth Rate) to two years, explained in Annex III of the ESTAT SDGs annual report for 2021. CAGR=((lastyearvalue/firstyearvalue)^(1/number of years))-1. We apply this to the dataset, so we compute a new column with CAGR for each MS,then we compute the mean for the EU based on these numbers.
```{r setup, include=FALSE} 
#knitr::opts_chunk$set(warning = FALSE, message = FALSE) 


# process CAGR for MS and EU-27 
pcchange_EU_MS <-merged_EU_MS %>%
    group_by(indicator_id, type, geo_code) %>% 
    arrange(time, .by_group = TRUE) %>%
    mutate(CAGR = (((value/lag(value))**(1/6))-1)*100) %>% 
    na.omit %>%
    select(-value) %>%  
    subset(geo_code!="EU_V")

##add in ecosystem condition indicators manually (to be published at end of 2021)
extra<-read.csv("ecosystemconditionindicators.csv")
merged_subf_esc <-bind_rows(pcchange_EU_MS,extra)
pcchange_EU_MS<-merged_subf_esc %>%
  subset(geo_code=="EU27_2020")%>% #if we want to isolate only EU27 - but then winorisation is not possible
  arrange(indicator_id)%>%
  distinct(geo_code, type, indicator_id,time,.keep_all = TRUE)
```
### Reshape dataframe from long to wide
Reshape the dataframe to a wide format as this is the format expected by COINr package.

```{r}
#convert from long to wide format for COINr tooland to make graphs of indicator progress
pcchange_EU_MS$indicator <- paste(pcchange_EU_MS$indicator_id, "_",pcchange_EU_MS$type)
merged_subf_ind = subset(pcchange_EU_MS, select = -c(indicator_id,type) )
forplot<-merged_subf_ind
forplot$IndCode<-forplot$indicator
joinforplot<-plyr::join(forplot,metadata,by="IndCode" )%>%
  mutate(Direction*CAGR) %>%
  select(-indicator,-IndName,-IntUnit)
write.csv(joinforplot,"joinforplot.csv",quote = FALSE, row.names=FALSE)
### Plot rate of change for each Key component in sunburst

# Add lines to the initial dataset
joinforplotman<-read.csv("joinforplot_man.csv")
data <- joinforplotman

kcncobj<-joinforplotman %>%
  group_by(Agg_NC)%>%
  mutate(meannc=mean(CAGRcorrdir)) %>%
  group_by(Agg_KC)%>%
  mutate(meankc=mean(CAGRcorrdir)) %>%
  group_by(Agg_OBJ)%>%
  mutate(meanobj=mean(CAGRcorrdir)) 

write.csv(kcncobj,"summarybynckcobj.csv")

#CIRCULAR BAR
empty_bar<-10
to_add <- data.frame( matrix(NA, empty_bar*nlevels(data$Agg_OBJ), ncol(data)) )
colnames(to_add) <- colnames(data)
to_add$Agg_OBJ <- rep(levels(data$Agg_OBJ), each=empty_bar)
data <- rbind(data, to_add)
data <- data %>% arrange(Agg_OBJ)
data$id <- seq(1, nrow(data))

label_data <- data
number_of_bar <- nrow(label_data)
angle <- 90 - 360 * (label_data$id-0.5) /number_of_bar     # I substract 0.5 because the letter must have the angle of the center of the bars. Not extreme right(1) or extreme left (0)
label_data$hjust <- ifelse( angle < -90, 1, 0)
label_data$angle <- ifelse(angle < -90, angle+180, angle)

# prepare a data frame for base lines
base_data <- data %>% 
  group_by(Agg_OBJ) %>% 
  summarize(start=min(id), end=max(id) - empty_bar) %>% 
  rowwise() %>% 
  mutate(title=mean(c(start, end)))
 
# prepare a data frame for grid (scales)
grid_data <- base_data
grid_data$end <- grid_data$end[ c( nrow(grid_data), 1:nrow(grid_data)-1)] + 1
grid_data$start <- grid_data$start - 1
grid_data <- grid_data[-1,]

p <- ggplot(data, aes(x=as.factor(id), y=CAGRcorrdir, fill=Agg_OBJ)) +       # Note that id is a factor. If x is numeric, there is some space between the first bar
  
  geom_bar(aes(x=as.factor(id), y=CAGRcorrdir, fill=Agg_OBJ), stat="identity", alpha=0.5) +
  
  # Add a val=100/75/50/25 lines. I do it at the beginning to make sur barplots are OVER it.
  geom_segment(data=grid_data, aes(x = end, y = 100, xend = start, yend = 100), colour = "grey", alpha=1, size=0.3 , inherit.aes = FALSE ) +
  geom_segment(data=grid_data, aes(x = end, y = 80, xend = start, yend = 80), colour = "grey", alpha=1, size=0.3 , inherit.aes = FALSE ) +
  geom_segment(data=grid_data, aes(x = end, y = 60, xend = start, yend = 60), colour = "grey", alpha=1, size=0.3 , inherit.aes = FALSE ) +
  geom_segment(data=grid_data, aes(x = end, y = 40, xend = start, yend = 40), colour = "grey", alpha=1, size=0.3 , inherit.aes = FALSE ) +
  geom_segment(data=grid_data, aes(x = end, y = 20, xend = start, yend = 20), colour = "grey", alpha=1, size=0.3 , inherit.aes = FALSE ) +
  # Add text showing the CAGRcorrdir of each 100/75/50/25 lines
  annotate("text", x = rep(max(data$id),4), y = c(20, 40, 60, 80,100), label = c("20", "40", "60", "80","100") , color="grey", size=3 , angle=0, fontface="bold", hjust=1) +
  
  geom_bar(aes(x=as.factor(id), y=CAGRcorrdir, fill=Agg_OBJ), stat="identity", alpha=0.5) +
  ylim(-100,120) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    plot.margin = unit(rep(-1,4), "cm") 
  ) +
  coord_polar() + 
  geom_text(data=label_data, aes(x=id, y=CAGRcorrdir+10, label=IndName, hjust=hjust), color="black", fontface="bold",alpha=0.6, size=2.5, angle= label_data$angle, inherit.aes = FALSE ) +
  
  # Add base line information
  geom_segment(data=base_data, aes(x = start, y = -5, xend = end, yend = -5), colour = "black", alpha=0.8, size=0.6 , inherit.aes = FALSE )  +
  geom_text(data=base_data, aes(x = title, y = -18, label=Agg_OBJ), hjust=c(1,1,0,0), colour = "black", alpha=0.8, size=4, fontface="bold", inherit.aes = FALSE)
 
p

kc<-joinforplotman %>%
  group_by(Agg_KC)%>%
  mutate(meankc=mean(CAGRcorrdir)) 
data <- kc 
empty_bar<-10
to_add <- data.frame( matrix(NA, empty_bar*nlevels(data$Agg_OBJ), ncol(data)) )
colnames(to_add) <- colnames(data)
to_add$Agg_OBJ <- rep(levels(data$Agg_OBJ), each=empty_bar)
data <- rbind(data, to_add)
data <- data %>% arrange(Agg_OBJ)
data$id <- seq(1, nrow(data))

label_data <- data
number_of_bar <- nrow(label_data)
angle <- 90 - 360 * (label_data$id-0.5) /number_of_bar     # I subtract 0.5 because the letter must have the angle of the center of the bars. Not extreme right(1) or extreme left (0)
label_data$hjust <- ifelse( angle < -90, 1, 0)
label_data$angle <- ifelse(angle < -90, angle+180, angle)

p <- ggplot(data, aes(x=as.factor(id), y=meankc, fill=Agg_OBJ)) +       # Note that id is a factor. If x is numeric, there is some space between the first bar
  geom_bar(stat="identity", alpha=0.5) +
  ylim(-15,15) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    plot.margin = unit(rep(-1,4), "cm") 
  ) +
  coord_polar() + 
  geom_text(data=label_data, aes(x=id, y=meankc+10, label=Agg_OBJ, hjust=hjust), color="black", fontface="bold",alpha=0.6, size=2.5, angle= label_data$angle, inherit.aes = FALSE ) 
 
p 


nc<-joinforplotman %>%
  group_by(Agg_NC)%>%
  mutate(meannc=mean(CAGRcorrdir)) 

write.csv(nc,"summarybynckcobj.csv")
data <- nc 
empty_bar<-4
to_add <- data.frame( matrix(NA, empty_bar*nlevels(data$Agg_OBJ), ncol(data)) )
colnames(to_add) <- colnames(data)
to_add$Agg_OBJ <- rep(levels(data$Agg_OBJ), each=empty_bar)
data <- rbind(data, to_add)
data <- data %>% arrange(Agg_OBJ)
data$id <- seq(1, nrow(data))

label_data <- data
number_of_bar <- nrow(label_data)
angle <- 90 - 360 * (label_data$id-0.5) /number_of_bar     # I subtract 0.5 because the letter must have the angle of the center of the bars. Not extreme right(1) or extreme left (0)
label_data$hjust <- ifelse( angle < -90, 1, 0)
label_data$angle <- ifelse(angle < -90, angle+180, angle)

p <- ggplot(data, aes(x=as.factor(id), y=meannc, fill=Agg_OBJ)) +       # Note that id is a factor. If x is numeric, there is some space between the first bar
  geom_bar(stat="identity", alpha=0.5) +
  ylim(-15,15) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    plot.margin = unit(rep(-1,4), "cm") 
  ) +
  coord_polar() + 
  geom_text(data=label_data, aes(x=id, y=meannc+10, label=Agg_NC, hjust=hjust), color="black", fontface="bold",alpha=0.6, size=2.5, angle= label_data$angle, inherit.aes = FALSE ) 
 
p 

```
```{r}
coininput<-merged_subf_ind %>%
  pivot_wider(names_from = indicator, values_from = CAGR) %>%
  mutate(UnitName=geo_code, UnitCode=geo_code) %>%
  mutate(Year=time) %>%
  subset(geo_code!="HG")%>%
  na_if("Inf")%>%
  relocate(UnitName,UnitCode, Year) 
coininput_sub = subset(coininput, select = -c(geo_code, time) )
write.csv(coininput_sub,"coininputf.csv",quote = FALSE, row.names=FALSE)

```
Read in the structural hierarchy of the EU Bioeconomy Monitoring System
```{r}
metadata<-read.csv("metadata_.csv")
AggregateMeta<-read.csv("AggregateMeta.csv")
```
To check whether or not the indicators are correctly mapped to their parent groups, we generate a figure showing the hierarchy based on the data.

```{r}
# assemble indicators, metadata and aggregate metadata files#################################################
indicators<- coininput_sub 
assembled_ALL <- assemble(indicators, metadata, AggregateMeta) |>
  suppressMessages()
assembled_ALL <- impute(assembled_ALL, dset = "Raw", imtype = "ind_mean")
assembled_ALL <- aggregate(assembled_ALL, dset = "Imputed",out2 = "df")
write.csv(assembled_ALL,"assembled_ALL2017.csv",quote = FALSE, row.names=FALSE)

wide<-assembled_ALL
assembled_ALL <- wide%>%
  subset(UnitName=="EU27_2020")
long <- assembled_ALL %>% 
  gather(key="indicator", value= "CAGR",-c(Year)) %>%
  mutate(CAGR2= as.numeric(CAGR,length=2))
long = long[-c(1:56),]

# plot framework
plotframework(assembled_ALL)
##############################################################################################################
```
## Processing indicators 

### Statistics of raw data
To investigate the data better, we have a look at the statistics.
```{r}
# get stats
assembled_ALL <- getStats(assembled_ALL, dset = "Raw")
# display in table using Reactable
assembled_ALL$Analysis$Raw$StatTable |>
  roundDF() |>
  reactable::reactable(resizable = TRUE, bordered = TRUE,
                       highlight = TRUE, defaultPageSize = 10)

```

The missing data points are identified.

```{r}
assembled_ALL <- checkData(assembled_ALL, dset = "Raw")
head(assembled_ALL$Analysis$Raw$MissDatSummary)
```

### Imputation
```{r}
# impute using indicator mean

assembled_ALL <- impute(assembled_ALL, dset = "Raw", imtype = "ind_mean")
# check how many NAs after imputation
head(sum(is.na(assembled_ALL$Data$Imputed)))


```
### Data treatment for outliers 
```{r,fig.cap = "Example of distribution of indicator 3.1.a.2-Material footprint (biomass) per capita before and after outlier treatment"}
# treat using defaults
assembled_ALL <- treat(assembled_ALL, dset = "Imputed", winmax = 5,deflog = "CTlog")
#check results
head(assembled_ALL$Analysis$Treated$TreatSummary %>% filter(Treatment != "None"))
iplotIndDist2(assembled_ALL, dsets = c("Imputed", "Treated"), icodes = "3.1.a.2 _ Material_Footprint_(Biomass)_per_capita",
              ptype = "Histogram")
#NORMALISE
assembled_ALL <- normalise(assembled_ALL, dset="Imputed", ntype="goalposts", npara = list(goalposts = c(-10,10,1)))
assembled_ALL <- aggregate(assembled_ALL, dset = "Imputed",out2 = "df")

wide<-assembled_ALL
write.csv(assembled_ALL,"assembled_ALL2017.csv",quote = FALSE, row.names=FALSE)
```

## Results
### Compare EU-27_2020 results to MS
```{r}
indicators<- coininput_sub 
assembled_ALL <- assemble(IndData = indicators, IndMeta = metadata, AggMeta = AggregateMeta)|>
  suppressMessages()
assembled_ALL <- impute(assembled_ALL, dset = "Raw", imtype = "ind_mean")
assembled_ALL <- treat(assembled_ALL, dset = "Imputed", winmax = 5)
assembled_ALL <- aggregate(assembled_ALL, dset = "Treated")
iplotBar(assembled_ALL, dset = "Aggregated", isel = "Food", usel = "EU-27", aglev = 4)
```
```{r}
indicators<- coininput_sub 
assembled_ALL <- assemble(IndData = indicators, IndMeta = metadata, AggMeta = AggregateMeta)|>
  suppressMessages()
assembled_ALL <- impute(assembled_ALL, dset = "Raw", imtype = "ind_mean")
assembled_ALL <- treat(assembled_ALL, dset = "Imputed", winmax = 5)
assembled_ALL <- aggregate(assembled_ALL, dset = "Treated")
iplotBar(assembled_ALL, dset = "Aggregated", isel = "Resources", usel = "EU-27", aglev = 4)
```
```{r}
indicators<- coininput_sub 
assembled_ALL <- assemble(IndData = indicators, IndMeta = metadata, AggMeta = AggregateMeta)|>
  suppressMessages()
assembled_ALL <- impute(assembled_ALL, dset = "Raw", imtype = "ind_mean")
assembled_ALL <- treat(assembled_ALL, dset = "Imputed", winmax = 5)
assembled_ALL <- aggregate(assembled_ALL, dset = "Treated")
iplotBar(assembled_ALL, dset = "Aggregated", isel = "Defossilization", usel = "EU-27", aglev = 4)
```
```{r}
assembled_ALL <- assemble(IndData = indicators, IndMeta = metadata, AggMeta = AggregateMeta)|>
  suppressMessages()
assembled_ALL <- impute(assembled_ALL, dset = "Raw", imtype = "ind_mean")
assembled_ALL <- treat(assembled_ALL, dset = "Imputed", winmax = 5)
assembled_ALL <- aggregate(assembled_ALL, dset = "Treated")
iplotBar(assembled_ALL, dset = "Aggregated", isel = "Climate", usel = "EU-27", aglev = 4)
```
```{r}
assembled_ALL <- assemble(IndData = indicators, IndMeta = metadata, AggMeta = AggregateMeta)|>
  suppressMessages()
assembled_ALL <- impute(assembled_ALL, dset = "Raw", imtype = "ind_mean")
assembled_ALL <- treat(assembled_ALL, dset = "Imputed", winmax = 5)
assembled_ALL <- aggregate(assembled_ALL, dset = "Treated")
iplotBar(assembled_ALL, dset = "Aggregated", isel = "Competitiveness", usel = "EU-27", aglev = 4)
```
```{r}
#write to long format
assembled_ALL <- wide%>%
  subset(UnitName=="EU-27")
long <- assembled_ALL %>% 
  gather(key="indicator", value= "CAGR",-c(Year)) %>%
  mutate(CAGR2= as.numeric(CAGR,length=2))
long = long[-c(1:56),]
```


## Plots
###Plot rate of change for Strategy Objectives
```{r, fig.cap="Compound annual growth rate between 2012-2017 per Strategy Objective"}
final_hier<-long%>%
  mutate(Code = indicator)%>%
  select(-Year) %>%
  merge(AggregateMeta, by = "Code", all = TRUE)%>%
  na.omit()%>%
  select(-indicator,-Weight) %>%
  subset(AgLevel==4)
parents=list(final_hier$Name)
ggplot(final_hier, aes(x=Code,y=round(as.numeric(CAGR), digits = 2))) +
  geom_col() + 
  coord_flip()
```
### Plot rate of change for Normative criteria level
```{r, fig.cap="Average annual rate of change between 2012-2017 per normative criteria"}
final_hier<-long%>%
  mutate(Code = indicator)%>%
  select(-Year) %>%
  merge(AggregateMeta, by = "Code", all = TRUE)%>%
  na.omit()%>%
  select(-indicator,-Weight) %>%
  subset(AgLevel==3)
parents=list(final_hier$Name)
ggplot(final_hier, aes(x=Code,y=round(as.numeric(CAGR), digits = 2))) +
  geom_col() + 
  coord_flip()
```
### Plot rate of change for each Key component
```{r, fig.cap="Average annual rate of change between 2012-2017 per Key component"}
final_hier<-long%>%
  mutate(Code = indicator)%>%
  select(-Year) %>%
  merge(AggregateMeta, by = "Code", all = TRUE)%>%
  na.omit()%>%
  select(-indicator,-Weight) %>%
  subset(AgLevel==2)
parents=list(final_hier$Name)
ggplot(final_hier, aes(x=Code,y=round(as.numeric(CAGR), digits = 2))) +
  geom_col() + 
  coord_flip()
```

